Gray Box Optimization using the Parameter-less Population Pyramid

### 2015-01-09 Gray Box Optimization using the Parameter-less Population Pyramid/main.tex
More Gray Box

P3 combines hill climbing and model building using a novel, iteratively constructed, pyramid
of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
works. In order to perform solution mixing, P3 learns a linkage tree from pairwise variable
entropy which is then used to perform global optima mixing~\cite{thierens:2011:gomea}. In global
optimal mixing a series of variable subsets are used to copy genetic information from multiple
sources into a single solution. Each time variables are donated, the modified solution is evaluated,
with the change being reverted if solution quality did not improve. In this way P3 performs
multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
donors in the pyramid level.
Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
memory usage.

@@

TODO Efficiency comparison of cluster creation / usage, creates new clusters every time

TODO Random subsample of clusters, not smallest first

TODO Doesn't recalculate entire fitness, just tracks delta changes.

TODO State that unlike BBP3, GBP3 performs hill climbing after each donation

TODO Compare P3 storage with TUX to highlight full population versus single solution mixing

>>>>

P3 combines hill climbing and model building using a novel, iteratively constructed, pyramid
of populations. Figure~\ref{fig-p3} provides the high level description of how the algorithm
works. In order to perform solution mixing, P3 learns a linkage tree from pairwise variable
entropy which is then used to perform global optima mixing~\cite{thierens:2011:gomea}. In global
optimal mixing a series of variable subsets are used to copy genetic information from multiple
sources into a single solution. Each time variables are donated, the modified solution is evaluated,
with the change being reverted if solution quality did not improve. In this way P3 performs
multiple evaluations per mixing event, with a single solution receiving genes from randomly chosen
donors in the pyramid level. This also means that unlike TUX, which uses information from only two
solutions to perform crossover, P3 uses information from an entire population.
Due to this model building, P3 requires \BigO{N} amortized time per evaluation and \BigO{N^2}
memory usage.

@@

This method of constructing clusters has a number of desirable properties beyond its
efficiency. As all non-linear relationships are denoted by subfunctions, mixing
complete subfunctions between individual has a high likelihood of preserving quality.
Each cluster created contains at least one entire subfunction, and two
clusters must differ by at least one entire subfunction. The probability two
variables appear in the same cluster increases with the number of subfunctions
they share. The more paths that exist between two variables in the subfunction
connectivity graph, the more likely those variables will appear in the same cluster.

Due to this change, a few compensatory adjustments to P3's algorithm were required.
First, in black box P3 the number of clusters discovered by linkage learning is
limited by the number of solutions at that level of the pyramid. To duplicate this
behavior, after calling \Call{SubfunctionTree}{} a random subset of the discovered
clusters is chosen, such that the size is limited to the number of solutions
stored at the current level of the pyramid. As the clusters produced do not
rely on the solutions stored in the pyramid, clusters are regenerated before each
mixing event, unlike black box P3 in which they are regenerated only when the stored
solutions change. Finally, unlike previous work, the clusters are applied in random order
to avoid the time required to sort them based on length.

Another efficiency gain possible due to the gray box domain is that each time a
donation is made, only the effected part of the solution's fitness needs to be
recalculated. As a result the number of subfunction evaluations required
to determine the change in fitness is only \BigO{|cluster|}. This also
allows for gray box P3 to efficiently reapply hill climbing after each
donation as only effected moves need to be rechecked. As a result a donation,
and its resulting modifications from hill climbing, are kept only if the
new local optima is at least as fit as the solution before the donation occurred.
All combined a single donation plus returning to a local optimum requires \BigO{|cluster| + I} time,
while just the donation in black box P3 requires \BigO{N}.
